---
title: "Pollution in Australia by Postcode"
author: "Neil Saunders"
date: "compiled `r Sys.time()`"
output: 
  github_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(tidyverse)
library(rvest)
library(rgdal)
library(pander)

theme_set(theme_minimal())
```

# Introduction
In this document we extract the data from the ABC News article [Australia's pollution mapped by postcode reveals nation's 'dirty truth'](https://www.abc.net.au/news/2018-11-16/australian-pollution-mapped-by-postcodes/10478620) and visualise it on maps.

## Getting the pollution data
Very easy using `rvest`. It even converts numeric strings with commas to numeric values.

We make a few changes:

- rename the multi-word column headers
- add a new column with zero-padded postcode, to join later
- convert the "Level" description to a correctly-ordered factor

```{r get-abc-data}
auspollution <- read_html("https://www.abc.net.au/news/2018-11-16/australian-pollution-mapped-by-postcodes/10478620") %>% 
  html_node("table") %>% 
  html_table() %>% 
  rename(facilities = `Number of polluting facilities`,
         emissions = `Total emissions in the air (kg)`,
         level = `Level of pollution`) %>% 
  mutate(POA_CODE16 = sprintf("%04d", Postcode),
         level = factor(level, levels = c("No polluting facilities", "Low", "Medium", "High")))

auspollution %>% 
  head(10) %>% 
  pander(justify = c("right", "right", "right", "left", "right"))
```

## Getting a shapefile of postcodes
You can download the ZIP file _Postal Areas ASGS Ed 2016 Digital Boundaries in ESRI Shapefile Format_ from [this ABS page](http://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/1270.0.55.003July%202016?OpenDocument), unzip and retain just the `.shp` file.

Then read it in:

```{r read-shapefile}
# seems to need hard-coded full path :(
auspost <- readOGR("../../data/", "POA_2016_AUST")
```

## Joining the spatial and pollution data
We can convert the shapefile data to a data frame using `fortify`, then join the `id` column with the `POA_CODE16` column created previously:

```{r join-data}
auspost_f <- fortify(auspost, region = "POA_CODE16") %>% 
  left_join(auspollution, by = c("id" = "POA_CODE16"))
```

## Metro area postcodes
We don't want to plot the entire country. [This handy PDF](http://www.impactlists.com.au/ImpactLists/media/list-tools/Useful-Postcode-Ranges.pdf) includes postcode ranges for Australian cities. As examples we'll use Sydney, Melbourne and Brisbane.

```{r metro-postcodes}
sydney <- as.character(c(1000:2249, 2760:2770))
melbourne <- as.character(c(3000:3207, 8000:8499))
brisbane <- as.character(c(4000:4207, 9000:9499))
```

# The maps

## Total emissions in the air (kg) - Sydney
The range of values is such that most postcodes have low values that are difficult to distinguish, even with the viridis colour scale.

The Botany area stands out.

```{r map-sydney}
auspost_f %>% 
  filter(id %in% sydney) %>% 
  ggplot(aes(long, lat, group = group)) +
    geom_polygon(aes(fill = emissions), color = "grey70") +
    coord_map() +
    scale_fill_viridis_c() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          panel.grid = element_blank()) +
    labs(title = "Total emissions in the air (kg) by Sydney postcode")
```

##  Level of pollution - Melbourne
Using the discrete values for pollution level results in arguably a more interesting map.

```{r map-melbourne}
auspost_f %>% 
  filter(id %in% melbourne) %>% 
  ggplot(aes(long, lat, group = group)) +
    geom_polygon(aes(fill = level), color = "grey70") +
    coord_map() +
    scale_fill_viridis_d() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          panel.grid = element_blank()) +
    labs(title = "Level of pollution by Melbourne postcode")
```

## Number of polluting facilities - Brisbane
The number of facilities provides a better continuous range than emissions. We see most facilities where the river meets Moreton Bay (including some of the islands), and south of the city.

```{r map-brisbane}
auspost_f %>% 
  filter(id %in% brisbane) %>% 
  ggplot(aes(long, lat, group = group)) +
    geom_polygon(aes(fill = facilities), color = "grey70") +
    coord_map() +
    scale_fill_viridis_c() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          panel.grid = element_blank()) +
    labs(title = "Number of polluting facilities by Brisbane postcode")
```
