---
title: "Diagnostic Test Accuracy"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    orientation: rows
runtime: shiny
---

```{r global, include=FALSE}
# load data in 'global' chunk so it can be shared by all users of the dashboard
library(waffle)
library(binom)
library(kableExtra)
```

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
sliderInput("prev", label = "Prevalence",
            min = 0, max = 1, value = 0.05, step = 0.01)

sliderInput("sens", label = "Sensitivity:",
            min = 0, max = 1, value = 0.95, step = 0.01)

sliderInput("spec", label = "Specificity:",
            min = 0, max = 1, value = 0.95, step = 0.01)
```

Row
-----------------------------------------------------------------------

### **Statistics**

```{r}
renderUI({
  
  format_ci <- function(ci) {
  lower <- sprintf("%.03f", ci$lower)
  upper <- sprintf("%.03f", ci$upper)
  
  paste(lower, upper, sep = " - ")
}

  tp <- round(input$prev * 1000 * input$sens)
  fn <- round(input$prev * 1000 - tp)
  tn <- round((1000 - input$prev * 1000) * input$spec)
  fp <- round((1000 - input$prev * 1000) - tn)
  
  sens_ci <- binom.confint(tp, tp + fn, methods = "exact")
  spec_ci <- binom.confint(tn, tn + fp, methods = "exact")
  acc_ci  <- binom.confint(tp + tn, tp + fp + tn + fn, methods = "exact")
  
  ppv_ci <- binom.confint(tp, tp + fp, methods = "logit")
  npv_ci <- binom.confint(tn, tn + fn, methods = "logit")
  
  ppv <- (input$prev * input$sens) / ((input$sens * input$prev) +  (1 - input$spec) * (1 - input$prev))
  npv <- input$spec * (1 - input$prev) / ((1 - input$sens) * input$prev + input$spec * (1 - input$prev))
  plr <- input$sens / (1 - input$spec)
  nlr <- (1 - input$sens) / input$spec
  acc <- input$sens * input$prev + input$spec * (1 - input$prev)
  

  data01 <- data.frame(Statistic = c("Prevalence", "Sensitivity", "Specificity", "Positive Predictive Value", 
                                     "Negative Predictive Value", "Positive Likelihood Ratio", "Negative Likelihood Ratio", "Accuracy"),
                     Definition = c("percentage of population with the disease", 
                                    "probability that a test result will be positive when the disease is present", 
                                    "probability that a test result will be negative when the disease is not present", 
                                    "probability that the disease is present when the test is positive", 
                                    "probability that the disease is not present when the test is negative",
                                    "probability of a positive test given presence of disease / probability of a positive test given absence of disease", 
                                    "probability of a negative test given presence of disease / probability of a negative test given absence of disease", 
                                    "overall probability that a patient is correctly classified"),
                     Value = c(input$prev, 
                               input$sens, 
                               input$spec, 
                               sprintf("%1.3g", c(ppv, npv, plr, nlr, acc))),
                     `CI (95 %)` = c(NA, 
                            format_ci(sens_ci), 
                            format_ci(spec_ci), 
                            format_ci(ppv_ci), 
                            format_ci(npv_ci), 
                            paste(
                              sprintf("%.03f", exp(log(plr) - 1.96 * sqrt(1/tp - 1/(tp + fn) + 1/fp - 1/(fp + tn)))),
                              sprintf("%.03f", exp(log(plr) + 1.96 * sqrt(1/tp - 1/(tp + fn) + 1/fp - 1/(fp + tn)))),
                              sep = " - "),
                            paste(
                              sprintf("%.03f", exp(log(nlr) - 1.96 * sqrt(1/fn - 1/(fn + tp) + 1/tn - 1/(fp + tn)))),
                              sprintf("%.03f", exp(log(nlr) + 1.96 * sqrt(1/fn - 1/(fn + tp) + 1/tn - 1/(fp + tn)))),
                              sep = " - "),
                            format_ci(acc_ci)))
  
  data01 %>% 
    kable(col.names = c("Statistic", "Definition", "Value", "CI (95%)")) %>% 
    kable_styling(bootstrap_options = c("striped", "condensed")) %>% 
    HTML()
})

```


### **Test Results**

```{r}
renderPlot({
  tp <- round(input$prev * 1000 * input$sens)
  fn <- round(input$prev * 1000 - tp)
  tn <- round((1000 - input$prev * 1000) * input$spec)
  fp <- round((1000 - input$prev * 1000) - tn)
  
  waffle(c("True Negative" = tn, "True Positive" = tp, "False Negative" = fn, "False Positive" = fp), 
         rows = 25,
         legend_pos = "top",
         size = 1,
         title = "1000 people")
})
```


Row
-----------------------------------------------------------------------

### **Notes**

Given a disease prevalence, test sensitivity and test specificity, calculate a range of statistical measures and visualise the results.

**Confidence Intervals**

CI for sensitivity, specificity and accuracy is obtained from R `binom::binom.confint` using the exact (Pearson-Klopper) method.

CI for positive and negative predictive values is obtained from R `binom::binom.confint` using the logit method.

CI for postive and negative likelihood ratios is calculated using Altman's log method. See [this page](https://www2.ccrb.cuhk.edu.hk/stat/confidence%20interval/Diagnostic%20Statistic.htm) for the formulas.

**Resources**

This tool is based on the following resources.

1. [MEDCALC Diagnostic test evaluation calculator](https://www.medcalc.org/calc/diagnostic_test.php){target="_blank"}
1. [Diagnostic Test Calculator](http://araw.mede.uic.edu/cgi-bin/testcalc.pl){target="_blank"}
1. [C.I. Calculator: Diagnostic Statistics](https://www2.ccrb.cuhk.edu.hk/stat/confidence%20interval/Diagnostic%20Statistic.htm){target="_blank"}

For a clear, simple explanation of diagnostic test statistics try [False Positives and False Negatives](https://www.mathsisfun.com/data/probability-false-negatives-positives.html){target="_blank"}.


### **True Values**

```{r}
renderPlot({
  tp <- round(input$prev * 1000)
  tn <- round(1000 - input$prev * 1000)
  
  waffle(c("True Negative" = tn, "True Positive" = tp), 
         rows = 25,
         legend_pos = "top",
         size = 1,
         title = "1000 people")
})
```