---
title: "Nightingale's Rose Charts"
author: "Neil Saunders"
date: "compiled `r Sys.time()`"
output:
  github_document:
    toc: yes
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
always_allow_html: true
---

<style type="text/css">
.main-container {
  max-width: 80%;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      fig.path = "figures/")
library(tidyverse)
library(HistData)
library(knitr)
library(kableExtra)

theme_set(theme_bw())
```


# Summary
A brief exploration of Florence Nightingale's Crimean War dataset, inspired by "Florence Nightingale: Data Viz Pioneer", an episode of _Cautionary Tales_ [premiered by 99% Invisible](https://99percentinvisible.org/episode/florence-nightingale-data-viz-pioneer/). 

# The data
We can get the dataset `Nightingale` from the R package [histData](https://cran.r-project.org/web/packages/HistData/index.html).

It's quite small (24 rows) and looks like this:

```{r view-histdata}
Nightingale %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)
```

# Reshaping
The dataset is not tidy.

- each cause has its own column, rather than columns for cause + value
- columns are a mixture of rates and absolute values

We can select the rate columns and use `pivot_longer` to convert to long format.

```{r tidy-dataset}
Nightingale %>% 
  select(Date, Month, Year, contains("rate")) %>% 
  pivot_longer(cols = 4:6, names_to = "Cause", values_to = "Rate") %>% 
  mutate(Cause = gsub(".rate", "", Cause)) %>% 
  head() %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 10)
```

# Charts

The help page, `?Nightingale` provides some R code to generate polar area and line charts but it's somewhat dated and cumbersome. Let's give it the tidyverse treatment.

## The "rose chart"
The "rose chart", also called (incorrectly) a Coxcomb chart, or polar area chart, is a bar chart projected onto polar coordinates.

We can generate something very similar to Nightingale's original chart like this:

```{r rose-chart}
Nightingale %>% 
  select(Date, Month, Year, contains("rate")) %>% 
  pivot_longer(cols = 4:6, names_to = "Cause", values_to = "Rate") %>% 
  mutate(Cause = gsub(".rate", "", Cause),
         period = ifelse(Date <= as.Date("1855-03-01"), "April 1854 to March 1855", "April 1855 to March 1856"),
         Month = fct_relevel(Month, "Jul", "Aug", "Sep", "Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr", "May", "Jun")) %>% 
  ggplot(aes(Month, Rate)) + 
  geom_col(aes(fill = Cause), width = 1, position = "identity") + 
  coord_polar() + 
  facet_wrap(~period) +
  scale_fill_manual(values = c("skyblue3", "grey30", "firebrick")) +
  scale_y_sqrt() +
  theme_void() +
  theme(axis.text.x = element_text(size = 9),
        strip.text = element_text(size = 11),
        legend.position = "bottom",
        plot.background = element_rect(fill = alpha("cornsilk", 0.5)),
        plot.margin = unit(c(10, 10, 10, 10), "pt"),
        plot.title = element_text(vjust = 5)) +
  ggtitle("Diagram of the Causes of Mortality in the Army in the East")
```

## Column chart
We can't simply remove the polar coordinates, as this will place some months in the wrong position on the basic column chart. So now we use `Date` on the x-axis.

We can also indicate the period before the arrival of the Sanitary Commission using grey shading.

```{r column-chart}
Nightingale %>% 
  select(Date, Month, Year, contains("rate")) %>% 
  pivot_longer(cols = 4:6, names_to = "Cause", values_to = "Rate") %>% 
  mutate(Cause = gsub(".rate", "", Cause)) %>% 
  ggplot(aes(Date, Rate)) + 
  geom_rect(xmin = as.numeric(as.Date("1854-04-01")), xmax = as.numeric(as.Date("1855-03-01")), ymin = 0, ymax = Inf, fill = "grey90") +
  geom_col(aes(fill = Cause)) + 
  scale_fill_manual(values = c("skyblue3", "grey30", "firebrick")) +
  theme_void() +
  theme(axis.text.x = element_text(size = 9),
        strip.text = element_text(size = 11),
        legend.position = "bottom",
        plot.background = element_rect(fill = alpha("cornsilk", 0.5)),
        plot.margin = unit(c(10, 10, 10, 10), "pt"),
        plot.title = element_text(vjust = 5)) +
  scale_x_date(date_labels = "%b %Y") +
  ggtitle("Diagram of the Causes of Mortality in the Army in the East")
```

## Line chart

We can also show the data as a line chart.

```{r line-chart}
Nightingale %>% 
    select(Date, Month, Year, contains("rate")) %>% 
    pivot_longer(cols = 4:6, names_to = "Cause", values_to = "Rate") %>% 
    mutate(Cause = gsub(".rate", "", Cause)) %>% 
    ggplot(aes(Date, Rate)) + 
    geom_line(aes(color = Cause)) +
    geom_vline(xintercept = as.numeric(as.Date("1855-03-01")), linetype = "dashed") +
    scale_color_manual(values = c("skyblue3", "grey30", "firebrick")) +
    theme_void() +
    theme(axis.text.x = element_text(size = 9),
          strip.text = element_text(size = 11),
          legend.position = "bottom",
          plot.background = element_rect(fill = alpha("cornsilk", 0.5)),
          plot.margin = unit(c(10, 10, 10, 10), "pt"),
        plot.title = element_text(vjust = 5)) +
    scale_x_date(date_labels = "%b %Y") +
  ggtitle("Diagram of the Causes of Mortality in the Army in the East")
```

# Conclusions
The _Cautionary Tales_ podcast episode concludes that deaths from disease were falling before the arrival of the Sanitary Commission, and that this is obscured - perhaps deliberately - by the choice of the polar area chart.

It's a fair point. However, what we can't know is what would have happened through 1855 in the absence of the Sanitary Commission. Is there a hint of the same "double peak", with a seasonal cycle, but smaller? Is that evidence for the effect of sanitation improvement?
